---
title: "Fall 2025 / Preseason Predictors -> 2026 Season Outcomes: Pitching + Hitting"
author: "Brooke Bullis"
date: "October 4, 2026"
format: pdf
editor: visual
---

## What this does, and the headline finding up front

This builds **two** game-level panels, one for pitching and one for
hitting, per the "can I do both pitching and hitting?" direction:

- **Pitcher-game panel**: predictors are each pitcher's **fall 2025**
  (Sep 1 - Dec 31, 2025) Rapsodo bullpen/live-session summaries; the
  outcome is every individual 2026 game that pitcher threw. One row per
  pitcher-game.
- **Hitter-game panel**: predictors are each hitter's **widened preseason**
  (Sep 1, 2025 - Feb 12, 2026, i.e. anything strictly before the first 2026
  game on Feb 13) Rapsodo hitting-session summaries; the outcome is every
  individual 2026 game that hitter played. One row per hitter-game.

**Why two different windows?** The 4 roster pitchers (Reimers, Bautista,
Lopez, Springer) already have dense fall-2025 bullpen histories, so the
pitching side keeps the strict "fall" window. Hitting is different: almost
nobody's *earliest* tracked Rapsodo hitting session falls in fall 2025 --
most of the roster first shows up at the team's Jan 30 / Feb 6 / Feb 7 2026
preseason scrimmages. Under the strict fall-2025 window, only **2 of 16**
players had usable hitting predictor data; widening to "anything before the
first game" raises that to **14 of 16** -- a real, substantive improvement,
not a loophole. The 2 players still excluded from the hitter panel (Mia
Lopez, Audrey Reimers) have zero Rapsodo hitting sessions on file in either
window; both are already fully represented in the pitcher panel.

Sessions on/after Feb 13, 2026 (2/20 scrimmage, 4/1 live, etc.) are
excluded from both windows -- using in-season tracking data to "predict"
the season it came from would be data leakage.

Blast Motion's swing-tracking data isn't in either panel at all -- its
dashboard is an embedded panel that automation can't reach, and it's still
pending a raw export from Blast Motion / the team admin. If that export
ever arrives, it could plausibly fill in Lopez's and Reimers' hitting side.

```{r setup}
#| message: false
library(tidyverse)

rapsodo_dir <- "rapsodo_fall2025"
sagehens_dir <- "sagehens_2026"
```

## Load the Rapsodo sessions and the 2026 per-game logs

```{r load}
rapsodo_sessions <- read_csv(file.path(rapsodo_dir, "rapsodo_sessions_sep2025_oct2026.csv"))

pitching_games <- read_csv(file.path(sagehens_dir, "player_gamelog_pitching_2026.csv"))
batting_games  <- read_csv(file.path(sagehens_dir, "player_gamelog_batting_2026.csv"))
```

## Name crosswalk

Sage Hens lists players as "Last, First"; Rapsodo lists them as typed in by
whoever logged the session (inconsistent capitalization). Built by hand
from the two rosters collected this project -- double-check this before
relying on it further.

```{r crosswalk}
crosswalk <- tribble(
  ~player,              ~rapsodo_name,         ~is_pitcher,
  "Liu, Kylie",          "kylie liu",           FALSE,
  "Murguia, Natalie",    "natalie murguia",     FALSE,
  "Huryn, Sadie",        "Sadie Huryn",         FALSE,
  "Sefkow, Claire",      "claire sefkow",       FALSE,
  "Song, Yuri",          "Yuri Song",           FALSE,
  "Milton, Marie",       "Marie Milton",        FALSE,
  "Phillips, Gianna",    "Gianna Phillips",     FALSE,
  "Reimers, Audrey",     "Audrey Reimers",      TRUE,
  "Roberts, Sophia",     "Sophia Roberts",      FALSE,
  "Bautista, Ava",       "Ava Bautista",        TRUE,
  "Grande, Maya",        "Maya Grande",         FALSE,
  "Burnham, Violet",     "Violet Burnham",      FALSE,
  "Gibson, Summer",      "summer gibson",       FALSE,
  "Nguyen, Jiselle",     "Jiselle Nguyen",      FALSE,
  "Springer, Nyla",      "Nyla Springer",       TRUE,
  "Lopez, Mia",          "mia lopez",           TRUE,
)
norm_name <- function(x) str_squish(str_to_lower(x))
```

## Aggregate Rapsodo sessions per player/role, by window

```{r aggregate}
fall2025 <- rapsodo_sessions |>
  filter(in_fall2025_window) |>
  mutate(match_key = norm_name(player))

preseason <- rapsodo_sessions |>
  filter(in_preseason_window) |>
  mutate(match_key = norm_name(player))

summarise_sessions <- function(df) {
  df |>
    group_by(match_key, role) |>
    summarise(
      n_sessions = n(),
      metric1_avg = mean(metric1_value, na.rm = TRUE),
      metric2_avg = mean(metric2_value, na.rm = TRUE),
      earliest = min(date),
      latest = max(date),
      .groups = "drop"
    )
}

fall2025_summary <- summarise_sessions(fall2025)
preseason_summary <- summarise_sessions(preseason)

pitch_predictors <- fall2025_summary |>
  filter(role == "pitcher") |>
  select(match_key, fall2025_pitch_n = n_sessions,
         fall2025_pitch_avg_spin_rate = metric1_avg,
         fall2025_pitch_avg_velo = metric2_avg,
         fall2025_pitch_earliest = earliest, fall2025_pitch_latest = latest)

hit_predictors <- preseason_summary |>
  filter(role == "hitter") |>
  select(match_key, preseason_hit_n = n_sessions,
         preseason_hit_avg_distance = metric1_avg,
         preseason_hit_avg_exit_velo = metric2_avg,
         preseason_hit_earliest = earliest, preseason_hit_latest = latest)
```

## Pitcher-game panel

One row per pitcher-game: fall-2025 pitching predictors joined onto every
2026 game that pitcher threw.

```{r pitcher-panel}
pitcher_game_panel <- crosswalk |>
  filter(is_pitcher) |>
  mutate(match_key = norm_name(rapsodo_name)) |>
  inner_join(pitch_predictors, by = "match_key") |>
  inner_join(pitching_games, by = "player") |>
  select(-match_key, -rapsodo_name, -is_pitcher)

pitcher_game_panel
n_pitchers_usable <- n_distinct(pitcher_game_panel$player)
n_pitcher_games <- nrow(pitcher_game_panel)
```

## Hitter-game panel

One row per hitter-game: widened-preseason hitting predictors joined onto
every 2026 game that hitter played. `inner_join` on the predictors means a
player with zero Rapsodo hitting sessions (Lopez, Reimers) is naturally
excluded rather than silently given NA predictor rows.

```{r hitter-panel}
hitter_game_panel <- crosswalk |>
  mutate(match_key = norm_name(rapsodo_name)) |>
  inner_join(hit_predictors, by = "match_key") |>
  inner_join(batting_games, by = "player") |>
  select(-match_key, -rapsodo_name, -is_pitcher)

hitter_game_panel
n_hitters_usable <- n_distinct(hitter_game_panel$player)
n_hitter_games <- nrow(hitter_game_panel)

hitters_excluded <- crosswalk |>
  mutate(match_key = norm_name(rapsodo_name)) |>
  anti_join(hit_predictors, by = "match_key") |>
  pull(player)
hitters_excluded
```

## Diagnostics: how much better is the widened window?

```{r diagnostics}
tibble(
  panel = c("pitcher-game (fall-2025 window)", "hitter-game (widened preseason window)"),
  players_usable = c(n_pitchers_usable, n_hitters_usable),
  players_total = c(4, nrow(crosswalk)),
  total_game_rows = c(n_pitcher_games, n_hitter_games)
)

# for comparison: how many hitters would a STRICT fall-2025 window give?
n_hitters_fall2025_only <- fall2025_summary |>
  filter(role == "hitter") |>
  nrow()
n_hitters_fall2025_only
```

## Save

```{r save}
dir.create("model_dataset", showWarnings = FALSE)
write_csv(pitcher_game_panel, "model_dataset/pitcher_game_panel.csv")
write_csv(hitter_game_panel, "model_dataset/hitter_game_panel.csv")
```

## Notes for the Data Description / Limitations section

The headline change from the first pass: predicting 2026 outcomes from
preseason tracking data works noticeably better once "preseason" is
defined as *everything before the first game* rather than strictly *fall*.
That's a defensible redefinition, not a post-hoc fudge -- the team's own
practice calendar has most everyone showing up on Rapsodo for the first
time at the Jan 30 / Feb 6 / Feb 7 scrimmages, not in September. Two
genuine limitations remain worth stating plainly: (1) Mia Lopez and Audrey
Reimers have no Rapsodo hitting data in any window, so the hitter panel
necessarily leaves them out (they're still fully represented as pitchers);
(2) Blast Motion's fall/preseason swing-tracking data isn't in either panel
-- its dashboard couldn't be automated, and getting it requires a raw
export from Blast Motion or the team's admin, a separate follow-up outside
this script. Because each panel is now game-level rather than
player-level, the usual within-player correlation across a pitcher's or
hitter's games should be accounted for in any model fit on these panels
(e.g., player as a random effect, or cluster-robust standard errors) rather
than treating each game as fully independent.
